# Short intro to R

In Quarto documents like this one, we can write comments by just using plain text. In contrast, code needs to be within *code blocks*, like the one below. To execute a code block, you can click on the little "Play" button or press `Cmd/Ctrl + Shift + Enter` when your keyboard is hovering the code block. 

```{r}
2 + 2
```

That was our first R command, a simple math operation. Of course, we can also do more complex arithmetic:

```{r}
12345 ^ 2  / (200 + 25 - 6 * 2) # this is an inline comment, see the leading "#"
```

In order to *create* a code block, you can press `Cmd/Ctrl + Alt + i` or click on the little green "+C" icon on top of the script.

::: {.callout-tip}

## Exercise

Create your own code block below and run a math operation.

:::

### Objects

A huge part of R is working with *objects*. Let's see how they work:

```{r}
my_object <- 10
```

```{r}
my_object # to print the value of an object, just call its name
```

We can now use this object in our operations:

```{r}
2 ^ my_object
```

Or even create another object out of it:

```{r}
my_object2 <- my_object * 2
```

```{r}
my_object2
```





TODO: objects and common functions. Explain how to run Quarto code blocks, and how to render a document to html.
