# Calculus

In this section we'll focus on three big ideas from calculus: derivatives, optimization, and integrals.

## Derivatives

Derivatives are about (instantaneous) rate of change.

> "In the fall of 1972 President Nixon announced that the rate of increase of inflation was decreasing. This was the first time a sitting president used the third derivative to advance his case for reelection" ([Rossi 1996](https://www.ams.org/notices/199610/page2.pdf))

Let's dissect what Nixon might have said:

> Inflation's \[first derivative, of prices\] rate of increase \[second derivative\] is going down \[third derivative\].

A more graphical way to think about a derivatives is as a *slope*. Let's consider a linear function of the form $y = 2 x$:

```{r}
#| message: false
library(tidyverse) # could also just do library(ggplot2)
ggplot() +
  stat_function(fun = function(x){2 * x}, 
                xlim = c(-10, 10))
```

We can imagine any political variables in the x- and y-axes. What is the rate of change? In other words, what is the derivative? Remember that we can calculate the slope with:

$$
m = \frac{f(x_2) - f(x_1)}{x_2-x_1}
$$ 

Now consider another slightly more complicated function, a quadratic one, $y = x^2$:

```{r}
ggplot() +
  stat_function(fun = function(x){x ^ 2}, 
                xlim = c(-10, 10))
```

What happens when we apply our slope function?

::: callout-note
#### Exercise

1) Use the slope formula to calculate the rate of change between 5 and 6.

2) Use the slope formula to calculate the rate of change between 5 and 5.5.

3) Use the slope formula to calculate the rate of change between 5 and 5.1.
:::

Takeaway: here the derivative depends on the value of $x$. It is actually $2x$.

Differential calculus is about finding these derivatives in a more straightforward manner! We can generalize our slope formula as follows:

$$
m = \frac{f(x_1+ \Delta x) - f(x_1)}{\Delta x}
$$

The point is that when $\Delta x$ is arbitrarily small, we'll get our rate of change. Formalizing this:

$$
\lim_{\Delta x\to0} \frac{f(x_1+ \Delta x) - f(x_1)}{\Delta x} = \frac{d}{dx} f(x) = \frac{dy}{dx} = f'(x) 
$$

A few points on notation:

- $\frac{d}{dx} f(x)$ is read "The derivative of $f$ of $x$ with respect to $x$."
    + The variable with respect to which we're differentiating is the one that appears in the bottom (in the case above, this is $x$).

  ::: callout-warning
  While the above looks like a fraction, it's really not. Do not try to cancel out the $d$s!
  :::

- $f'(x)$ (read: "$f$ prime $x$") is the derivative of $f(x)$. This is a more compact form to refer to derivatives when you have defined $f(x)$ elsewhere. 


### Rules of differentiation

How to compute derivatives? Sometimes you can try a bunch of numbers and get at the answer. Sometimes you can use the limit-based formula above, if you know a few properties of limits. But in most cases you will either use software (more on this later) or the **rules of differentiation**, which we will cover now.

**Constant rule:** $(c)' = 0$.

There is no change in a constant:

```{r}
ggplot() +
  stat_function(fun = function(x){2}, xlim = c(-10, 10))
```

**Coefficient rule:** $(c \cdot f(x))' = c \cdot f'(x)$.

```{r}
ggplot() +
  stat_function(fun = function(x){2 * x}, xlim = c(-10, 10), aes(color = "y = 2x")) +
  stat_function(fun = function(x){4 * x}, xlim = c(-10, 10), aes(color = "y = 4x")) +
  scale_color_manual("Function", values = c("red", "blue"))
```

**Sum/difference rule:** $(f(x) \pm g(x))' = f'(x) \pm g'(x)$.

The two rules above give us that the derivative is a *linear operator*.

**Power rule:** $(x^n)'=nx^{(n-1)}$

Remember when we wanted to calculate the derivative of $y=x^2$ above? We can use the power rule, with $n=2$: $\;nx^{(n-1)} = 2x^{(2-1)}=2x$. Let's try out $\frac{d}{dx}4x^3$ and $\frac{d}{dx}(x^2 + 2x)$ on the board.

::: callout-note
#### Exercise

Use the differentiation rules we have covered so far to calculate the derivatives of $y$ with respect to $x$ of the following functions:

1) $y = 2x^2 + 10$
2) $y = 5x^4 - \frac{2}{3}x^3$
3) $y = 9 \sqrt x$
4) $y = \frac{4}{x^2}$
5) $y = ax^3 + b$, where $a$ and $b$ are constants.
6) $y = \frac{2w}{5}$
:::

**Exponent and logarithm rules:**

$$
\begin{aligned}
(c^x)' &= c^x \cdot ln(c), & \forall x>0 \\
(e^x)' &= e^x \\
\\
(log_a(x))' &= \frac{1}{x \cdot ln(a)}, & \forall x>0  \\
(ln(x))' &= \frac{1}{x}, & \forall x>0 
\end{aligned}
$$

We saw previously how Euler's number ($e$) arises from compound interest. The properties above make it very useful in a lot of calculus applications!

::: callout-note
#### Exercise

Compute the following:

1) $\frac{d}{dx}(10e^x)$
2) $\frac{d}{dx}(ln(x) - \frac{e^2}{3})$
:::

Now we'll get to a couple of more advanced (and powerful) rules.


**Product rule:** $(f(x)g(x))'=f'(x)g(x) + g'(x)f(x)$

Let's calculate $\frac{d}{dx}(3 \cdot ln(x) \cdot x^2)$ on the board.

**Quotient rule:** $\displaystyle(\frac{f(x)}{g(x)})' = \frac{f'(x)g(x) + g'(x)f(x)}{[g(x)]^2}$

**Chain rule:** $(f(g(x))' = f'(g(x)) \cdot g'(x)$

Let's compute $\frac{d}{dx}(e^{x^{2}})$ on the board.

::: callout-note
#### Exercise

Use the differentiation rules we have covered so far to calculate the derivatives of $y$ with respect to $x$ of the following functions:

1) $x^3 \cdot x$
2) $e^x \cdot x^2$
3) $(3x^4-8)^2$
:::

### Higher-order derivatives

We saw how politicians can refer to higher-order derivatives. To compute them, you simply "pass the outputs," starting from the lowest order and going up.

The second derivative tells us whether the slope of a function is increasing, decreasing, or staying the same at any point $x$ on the function's domain. For example, when driving a car:

-  $f(x)$ = distance traveled at time $x$
-  $f'(x)$ = speed at time $x$
-  $f''(x)$ = acceleration at time $x$

Let's compute the following second derivative:

$$f''(x^4) = \frac{d^2(x^4)}{dx^2}$$

- First, we take the first derivative: $f'(x^4)=4x^3$

- Then we use that output to take the second derivative: $f''(x^4)=f'(4x^3)=12x^2$

- We can keep going... for example, the third derivative: $$f'''(x^4) = f'(12x^2) = 24x$$

::: callout-note
#### Exercise

Compute the following:

1) $\frac{d^3}{dx^3}(x^5)$
2) $f''(4x^{3/2})$
3) $f''(4 \cdot ln(x))$
:::

### Partial derivatives

For a function $f(x,z)$, we might want to know how the function changes with respect to $x$. We call this a *partial derivative*:

$$ 
\frac{\partial}{\partial_x}f(x,z) = \frac{\partial_y}{\partial_x} = \partial_x f
$$

To obtain a partial derivative, we treat all other variables as constants and take the derivative with respect to the variable of interest (here $x$). For example:

$$
\begin{aligned}
y = f(x,z) &= xz \\
\frac{\partial_y}{\partial_x} &= z
\end{aligned}
$$

What is $\displaystyle\frac{\partial_y}{\partial_z}?$

Let's solve $\displaystyle\frac{\partial (x^2y+xy^2-x)}{\partial x}$ and $\displaystyle\frac{\partial (x^2y+xy^2-x)}{\partial y}$ on the board.

::: callout-tip
#### Example

Let's say that $y$ is how much I like a movie, $d$ is how many dogs a movie has, and $e$ is how many explosions a movie has. I claim that how much I like a movie can be expressed by a function of the type $y = f(d, e)$. Evaluate the following situations:

1. I like dogs and I don't care about action. So I believe that the true relationship is $y = f(d, e) = 3 \cdot d$. What is $\displaystyle\frac{\partial_y}{\partial_d}$, and how can we interpret it?

2. I like dogs and I like action. So I believe that the true relationship is $y = f(d, e) = 3 \cdot d + 1 \cdot e$. What is $\displaystyle\frac{\partial_y}{\partial_d}$, and how can we interpret it?

3. I like dogs and I like action. *But I definitely don't like them together*---I don't want the dogs to be in danger! So I believe that the true relationship is $y = f(d, e) = 3 \cdot d + 1 \cdot e -10 \cdot d \cdot e$. What is $\displaystyle\frac{\partial_y}{\partial_d}$, and how can we interpret it?

:::

::: callout-note
#### Exercise

Take the partial derivative with respect to $x$ and with respect to $z$ of the following functions. What would the notation for each look like?

1) $y = 3xz - x$
2) $x^3+z^3+x^4z^4$
3) $e^{xz}$
:::


### Differentiability of functions

Not all functions are differentiable at every point of their domains! 

An important concept here is whether functions are **continuous** at a point:

- Informally: A function is continuous at a point if its graph has no holes or breaks at that point

- Formally: A function is continuous at a point $a$ if: $\lim_{x \to a} f(x)=f(a)$

When is a function **differentiable** at a point?

-   If a function is differentiable at a point, it is also continuous at that point.

-   If a function is continuous at a point, it is *not* necessarily differentiable at that point.
    + Impossible to calculate derivative at sharp turns, cusps, or vertical tangents.

```{r}
ggplot() +
  stat_function(fun = function(x){abs(x) + 2}, xlim = c(-4, 4), 
                aes(color = "y = |x| + 2")) +
  stat_function(fun = function(x){sqrt(abs(x)) + 1}, xlim = c(-4, 4), 
                aes(color = "y = √(|x|) + 1")) +
  stat_function(fun = function(x){sign(x) * abs(x)^(1 / 3)}, xlim = c(-4, 4), 
                aes(color = "y = ₃√x")) +
  scale_colour_manual("Function", values = c("red", "blue", "black")) +
  labs(title = "Examples of functions that are not differentiable at x=0")
```


## Optimization

*Optimization* allows us to find the minimum or maximum values (or *extrema*) a function takes. It has many applications in the social sciences:

-   Formal theory: utility maximization, continuous choices
    
-   Ordinary Least Squares (OLS): Focuses on *minimizing* the squared errors between observed data and model-estimated values
    
-   Maximum Likelihood Estimation (MLE): Focuses on *maximizing* a likelihood function, given observed values.

### Identifying *extrema* analytically

On extrema: informally, a maximum is just the highest value a function takes, and a minimum is the lowest value.

In some situations, it can be easy to identify extrema intuitively by looking at a graph of the function (see the graphs above).

-   Maxima are high points ("peaks")
    
-   Minima are low points ("valleys")


We can use derivatives (rates of change!) to get at extrema.

-   When the derivative is zero (or fails to exist), the function has *usually* reached a (local) maximum or minimum.

    + At a maximum, the function must be increasing before the point and decreasing after it.

    + At a minimum, the function must be decreasing before the point and increasing after it.

-   We'll start by identifying points where this is the case ("critical points" or "stationary points").
  
::: callout-note
A point where $f'(x)=0$ or $f'(x)$ does not exist is called a *critical point* (or *stationary point*). Local extrema occur at critical points, but not all critical points are extrema. For instance, sometimes the graph is changing between concave and convex ("inflection points"). Or sometimes the function is not differentiable at that point for other reasons.
:::

We can find the local maxima and/or minima of a function by taking the derivative, setting it equal to zero, and solving for $x$ (or whatever variable). This gives us the first-order condition (FOC).

$$FOC: f'(x)=0$$

Notice that after this we only know that there is a critical point. **BUT** we don't know if we've found a maximum or minimum, or even if we've found an extremum.

To determine whether a we are seeing a (local) maximum or minimum, we can use the **Second Derivative Test**:

- Start by identifying $f''(x)$

- Substitute in the stationary points $(x^*)$ identified from the FOC.

    +   $f''(x^*) > 0$ we have a local minimum

    +   $f''(x^*) < 0$ we have a local maximum

    +   $f''(x^*) = 0$ we (may) have an inflection point - need to calculate higher-order derivatives (don't worry about this now)

Collectively these give use the **Second-Order Condition (SOC)**.

Now when it comes to knowing whether extrema are local or global:

-   To find the minimum/maximum on some interval, compare the local min/max to the value of the function at the interval's endpoints.

-   To find the global minimum/maximum, check the function's limits as it approaches $+ \infty$ and $- \infty$.

-   Extreme value theorem: if a real-valued function $f$ is continuous on the closed interval \[a,b\], then $f$ must attain a (global) maximum and a (global) minimum.


## Integrals

### Area under a curve

-   Often we want to find the area under a curve. 

-   Sometimes finding the area is easy. What's the area under the curve between $x=-1$ and $x=1$ for this function? $$f(x) = 
\begin{cases}
\frac{1}{3} & \text{for } x \in [0, 3] \\
0 & \text{otherwise}
\end{cases}$$

::: {.callout-tip title="\"Hint"}
We can draw this and look at the graph. Remember: $$area = \ell*w$$
:::

-   Normally, finding the area under a curve is much harder. But this is basically the question behind integration.

### Integrals as summation

-   We are already familiar with summation notation.
$$\displaystyle\sum_{i=1}^{n} i$$

-   This only works when we have discrete values to add. 

    -   When we need to add continuously, we have to use something else. Specifically, integrals.

### Definite integrals

-   Let's say we have a function $$ y = x^2 $$ And we want to find the area under the curve from $x=0$ to $x=1$. 

-   To find the area we're interested in here, we can use the definite integral.

-   Generally speaking, the notation looks like this:
$$\displaystyle\int_{x=a}^{b} f(x),dx$$
-   Here $a$ is the lower limit of integration, $b$ is the upper limit of integration, our function $f(x)$ is our integrand, and $x$ is our variable of integration.

------------------------------------------------------------------------

-   For our question, we're looking for the following: $$\displaystyle\int_{x=0}^{1} f(x) dx$$

-   This will give us a real number denoting the area under the curve of our function ($y=x^2$) between $x=0$ and $x=1$.

-   If $f$ is continuous on $[a,b]$ or bounded on $[a,b]$ with a finite number of discontinuities, then $f$ is integrable on $[a,b]$.

### Indefinite integrals

-   The *indefinite* integral, also known as the *anti-derivative*, $F(x)$ is the inverse of the function $f'(x)$. $$F(x)= \displaystyle\int f(x) \text{ } dx$$ 

-   This means if you take the derivative of $F(x)$, you wind up back at $f(x)$. $$F' = f \text{ or } \displaystyle\frac{dF(x)}{dx} = f(x)$$

-   This process is called anti-differentiation, or indefinite integration.

    -   While the definite integral gives us a real number (the total area under a curve), the indefinite integral gives us a function.

-   We need the concept of indefinite integrals to help us solve definite integrals.

### Solving definite integrals

-   The easiest way to calculate definite integrals, known as the "fundamental theorem of calculus," is shown below:

$$\displaystyle\int_{a}^{b} f(x) \text{ } dx = F(b)-F(a) = F(x)\bigg|_{a}^{b}$$

-   First we determine the antiderivative (indefinite integral) of $f(x)$ (and represent it $F(x)$), substitute the upper limit first and then the lower limit one by one, and subtract the results in order.

::: callout-warning
$C$ in the following definitions and rules is the called the "constant of integration." We need to add it when we define *all* antiderivatives (integrals) of a function because the anti-derivative "undoes" the derivative.

Remember that the derivative of any constant is zero. So if we find an integral $F(x)$ whose derivative is $f(x)$, adding (or subtracting) any constant will give us another integral $F(x)+C$ whose derivative is *also* $f(x)$.
:::

### Rules of integration

$$ \displaystyle\int_{a}^{a}f(x) \text{ }dx = 0$$

$$\displaystyle\int_{a}^{b} f(x) \text{ } dx = -\displaystyle\int_{b}^{a}f(x)dx$$

$$\int a \text{ }dx = ax + C \text { where $a$ is a constant}$$

$$\displaystyle\int af(x)dx = a\displaystyle\int f(x) \text{ }dx \text{ where $a$ is a constant}$$

$$\int (f(x) + g(x)) \text{ } dx = \int f(x) dx + \int g(x)dx$$

$$\int x^n dx = \frac{x^{n+1}}{n+1} + C \qquad \forall n \neq -1$$

$$\int x^{-1}dx = \ln |x| + C$$

### Solving the problem

Remember our function $y=x^2$ and our goal of finding the area under the curve from $x=0$ to $x=1$.

-   Find the indefinite integral, $F(x)$

    -   $\displaystyle\int x^2 \text{ } dx$
    
    -   $\displaystyle\frac{x^3}{3}+C$

:::{.callout-important}
We use the "power rule" of integration, which is the following: $$\int x^n dx = \frac{x^{n+1}}{n+1} + C \qquad \forall n \neq -1$$
:::

-   Evaluate at our lowest and highest points, $F(0)$ and $F(1)$.

    -   $F(0) = 0$
    
    -   $F(1) = \displaystyle\frac{1}{3}$
    
    -   Technically $0 + C$ and $\displaystyle\frac{1}{3} + C$, but the C's will fall out in the next step
    
-   Calculate $F(1) - F(0)$
$$\displaystyle\frac{1}{3} - 0 = \displaystyle\frac{1}{3}$$


::: callout-note
#### Exercise

Solve the following indefinite integrals:

1. $\int x^2 \text{ } dx$

2. $\int 3x^2\text{ } dx$

3. $\int x\text{ } dx$ 

4. $\int 3x^2 + 2x - 7\text{ }dx$

5. $\int \dfrac{2}{x}\text{ }dx$

And solve the following definite integrals:

1. $\displaystyle\int_{1}^{7} x^2 \text{ } dx$

2. $\displaystyle\int_{1}^{10} 3x^2 \text{ } dx$

3. $\int_7^7 x\text{ } dx$

4. $\displaystyle\int_{1}^{5} 3x^2 + 2x - 7\text{ }dx$ 

5. $\int_{1}^{e} \dfrac{2}{x}\text{ }dx$

:::
